Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2023-02-08 analysis/paper/paper.qmd
Before deleting "variation between sites"

-@eerkens2001techniques presented some average CV values (%) of artifacts made from various materials (See Table 1 from @eerkens2001techniques). 
Among those various material artifacts, the most comparable artifacts to the stemmed points are Great Basin projectiles points, whose CV value ranged from 6-55, with an average of 22. Also, projectile points from the US Southwest have CV values ranged from 11 to 33 [@garvey2018current]. Unlike correlation, @bettinger1999point had vaguely mention about the threshold for CV values. Therefore, here we sat `r threshold_value` as a rough threshold to evaluate the transmission biases referring to the example of Great Basin projectile points. We calculated CV values of Rosegate points from Monitor Valley, which was speculated as the byproducts of indirect bias (raw data from table 5 of @bettinger1999point). With the exception of weight, the CV range for metric attributes on the points is 17.01 to 24.59. We chose higher side on the CV range of Rosegate points as the threshold of our research based on following reasons. 1) Variation is generated by small errors that are transmitted between individuals and the errors gets bigger through generations [@eerkens2005cultural]. The duration of transmission process for stemmed points is much longer (approx.30,000yrs) than the case of Rosegate (approx.500yrs in Fremont region) [@bischoff2021rosegate]. 2) The errors are also highly associated with raw materials. Some raw materials, such as clay, are easier to control while, less predictable and controllable materials such as stone are likely to have inflated CV values [@eerkens2001techniques]. We assume that Rosegate projectile points were made from more finer grained raw materials such as flint compared to raw materials for stemmed points such as shale, rhyolite, porphyry, etc. (@fig-raw-materials-by-phase). 3) Compared to normal CV, corrected CV has slightly higher values [@vanpool2011quantitative]. Based on the threshold, we premised that CV values that are lower than `r threshold_value` represent guided variation while CV values over `r threshold_value` reflect indirect bias.

@@

## Correlation coefficients over time

To understand temporal patterns in the modes of cultural transmission, we grouped our assemblages into the three Korean Late Paleolithic chronological phases summarized in @tbl-korean-chronology, and computed correlation for all complete stemmed points each phase. Among the three choronological phases, we excluded Phase 1 from our analysis because there are only two complete stemmed points from Yonghodong and Sachang that date to this phase.

Our results show positive correlation between attributes throughout the Late Paleolithic. Panel A and B of @fig-corr-analysis indicate that some attributes such as BL and ML have stronger correlation (i.e. darker-blue points) than others (i.e. lighter-blue or almost invisible points). The correlations between attributes became stronger over time as we can clearly see in the the Panel C of @fig-corr-analysis. The median for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to evaluate transmission biases, while the Phase 3 median is close to the threshold. Based on this results, we assume that the earlier transmission for stemmed points was more likely the results of guided variation. On the other hand, Phase 3 could be more associated with indirect bias.

```{r}
#| label: fig-corr-analysis
#| fig-cap: "Correlation analysis by chronological phases in the Korean Late Paleolithic period. A: Correlation coefficient between attributes for Phase 1. B. Correlation coefficient between attributes for Phase 2. The individual point represesnts correlation between two attributes. Some points are invisible due to their weak relationship. C. Correlation coefficient for the second and third chronological phases. The grey points represent correlation between two attributes. "
#| fig-width: 7
#| fig-height: 3

@@

plot_grid(plot_grid(
          plot_grid(p2_corr_plot, 
                    p3_corr_plot,
                    ncol = 1,
                    labels = c("A", "B")),
          compare_p2_and_p3_correlations_plot,
          ncol = 2,
          labels = c("", "C"),
          rel_heights = c(1, 1.5)))

@@

```{r}
#| label: fig-plot-cor-four-assemblage
#| fig-cap: Correlation for assemblages with more than 5 stemmed points. Among the four assemblages, SYG1_2 and SYG6_2 are from Phase 2 of the Korean Late Paleolithic chronology and SYG1_3 and YS_3 are from Phase 3.
#| fig-width:   4.5
#| fig-height:  4

@@

One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points. These isolated finds are ambiguous with respect to a local tradition of artifact making, so to further investigate temporal change, we focus only the four assemblages that have five or more stemmed points in one chronological phase

```{r}
#| label: fig-corr-raw-ass
#| fig-cap: A. sth sth B. sth sth.

@@

correlations_by_site <- 
  ggplot(df_sitename_corr_site) +
  aes(sitename,
      value) +
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(size = 2,
                               alpha = 0.3) +
  theme_minimal() +
  labs(y = "Correlation coefficient",
       x = "Site") +
theme(axis.title.y = element_text(size = 7.5)) +
  # draw a line to show our threshold value
  geom_hline(yintercept = threshold_value_cor, colour = "red") +
  theme_bw()

@@

correlations_by_raw_material <- 
  ggplot(df_sitename_corr_raw_material) +
  aes(SPstage.Raw_material,
      value) +
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(size = 2,
                               alpha = 0.3) +
  theme_minimal() +
  labs(y = "Correlation coefficient",
       x = "Raw material") +
theme(axis.title.y = element_text(size = 7.5)) +
  # draw a line to show our threshold value
  geom_hline(yintercept = threshold_value_cor, colour = "red") +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))+
  theme_bw()

@@

## Coefficient of variation

@@

## Variation Over Time

@@

One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points. These isolated finds are ambiguous with respect to a local tradition of artifact making, so to further investigate temporal change, we focus only the four assemblages that have five or more stemmed points in one chronological phase (@fig-plot-cv-four-assemblage). These four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artifact making. The four assemblages include three from the Suyanggae site (SYG) and one from the Yongsandong site (YS). SYG1_2 and SYG6_2 are from Phase 2 (shown on the left side of the figure) and SYG1_3 and YS_3 are from Phase 3 (shown on the right side of the figure). Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width. @fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`. Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest ranges of confidence intervals, perhaps due to its small sample size, relative to the other assemblages compared here. Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-four-assemblage. This suggests that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.

>>>>

@eerkens2001techniques presented some average CV values (%) of artifacts made from various materials (See Table 1 from @eerkens2001techniques). 
Among those various material artifacts, the most comparable artifacts to the stemmed points are Great Basin projectiles points, whose CV value ranged from 6-55, with an average of 22. Also, projectile points from the US Southwest have CV values ranged from 11 to 33 [@garvey2018current]. Unlike correlation, @bettinger1999point had vaguely mentioned about the threshold for CV values. Therefore, here we sat `r threshold_value` as a rough threshold to evaluate the transmission biases referring to the example of Great Basin projectile points. We calculated CV values of Rosegate points from Monitor Valley, which was speculated as the byproducts of indirect bias (raw data from table 5 of @bettinger1999point). With the exception of weight, the CV range for metric attributes on the points is 17.01 to 24.59. We chose higher side on the CV range of Rosegate points as the threshold of our research based on following reasons. 1) Variation is generated by small errors that are transmitted between individuals and the errors gets bigger through generations [@eerkens2005cultural]. The duration of transmission process for stemmed points is much longer (approx.30,000yrs) than the case of Rosegate (approx.500yrs in Fremont region) [@bischoff2021rosegate]. 2) The errors are also highly associated with raw materials. Some raw materials, such as clay, are easier to control while, less predictable and controllable materials such as stone are likely to have inflated CV values [@eerkens2001techniques]. We assume that Rosegate projectile points were made from more finer grained raw materials such as flint compared to raw materials for stemmed points such as shale, rhyolite, porphyry, etc. (@fig-raw-materials-by-phase). 3) Compared to normal CV, corrected CV has slightly higher values [@vanpool2011quantitative]. Based on the threshold, we premised that CV values that are lower than `r threshold_value` represent guided variation while CV values over `r threshold_value` reflect indirect bias.

@@

## Correlation coefficients and Variation Over Time

Our results show positive correlation between attributes throughout the Late Paleolithic (Panel A of @fig-corr-analysis). Some relationships such as BL and ML have stronger correlation (i.e. darker-blue points) than others (i.e. lighter-blue or almost invisible points). To understand temporal patterns in the modes of cultural transmission, we grouped our assemblages into the three Korean Late Paleolithic chronological phases summarized in @tbl-korean-chronology. Among the three chronological phases, we excluded Phase 1 from our analysis because there are only two complete stemmed points from Yonghodong and Sachang that date to this phase.

The correlations between attributes became stronger over time as we can clearly see in the the Panel D of @fig-corr-analysis. The median for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to evaluate transmission biases, while the Phase 3 median is close to the threshold. Based on this results, we assume that the earlier transmission for stemmed points was more likely the results of guided variation. On the other hand, Phase 3 could be more associated with indirect bias.

```{r}
#| label: fig-corr-analysis
#| fig-cap: "Correlation analysis by chronological phases in the Korean Late Paleolithic period. A. Correlation coefficient between attributes for all periods. B. Correlation coefficient between attributes for Phase 1. C. Correlation coefficient between attributes for Phase 2. The individual point represesnts correlation between two attributes. Some points are invisible due to their weak relationship. D. Correlation coefficient for the second and third chronological phases. The grey points represent correlation between two attributes."
#| fig-width: 7
#| fig-height: 5

@@

plot_grid(plot_grid(
          plot_grid(
                    all_corr_plot,        
                    p2_corr_plot, 
                    p3_corr_plot,
                    ncol = 1,
                    labels = c("A", "B", "C")),
          compare_p2_and_p3_correlations_plot,
          ncol = 2,
          labels = c("", "D"),
          rel_heights = c(1, 2.5)))

@@

One limitation of this aggregation of all stemmed points in our sample into each chronological phase is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points. These isolated finds are ambiguous with respect to a local tradition of artifact making, so to further investigate temporal change, we focus only the four assemblages that have more than five stemmed points (@fig-plot-cor-four-assemblage). We assume these four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artifact making. The four assemblages include three from the Suyanggae site (SYG) and one from the Yongsandong site (YS). There are more variation in correlation coefficient in Phase 2 while the median values for the two assemblages from Phase 3 are similar to each other and close to `r threshold_value_cor`. Even though SYG1_2 is from Phase 2 but it has the higher median than other assemblages.

```{r}
#| label: fig-plot-cor-four-assemblage
#| fig-cap: Correlation for assemblages with more than 5 stemmed points. Among the four assemblages, SYG1_2 and SYG6_2 are from Phase 2 of the Korean Late Paleolithic chronology and SYG1_3 and YS_3 are from Phase 3. The grey points represent correlation between two attributes.
#| fig-width:   4.5
#| fig-height:  3.5

@@

```{r}
#| label: fig-corr-raw-ass
#| fig-cap: A. Correlation for each site that has more than five stemmed points. B. Correlation for raw materials that were used for making more than one stemmed poitns. The grey points represent correlation between two attributes.

@@

correlations_by_site <- 
  ggplot(df_sitename_corr_site) +
  aes(sitename,
      value) +
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(size = 2,
                               alpha = 0.3) +
  theme_minimal() +
  labs(y = "Correlation coefficient",
       x = "Site") +
theme(axis.title.y = element_text(size = 4)) +
  # draw a line to show our threshold value
  geom_hline(yintercept = threshold_value_cor, colour = "red") +
  theme_bw()

@@

correlations_by_raw_material <- 
  ggplot(df_sitename_corr_raw_material) +
  aes(SPstage.Raw_material,
      value) +
  geom_boxplot() +
  ggbeeswarm::geom_quasirandom(size = 2,
                               alpha = 0.3) +
  theme_minimal() +
  labs(y = "Correlation coefficient",
       x = "Raw material") +
theme(axis.title.y = element_text(size = 4)) +
  # draw a line to show our threshold value
  geom_hline(yintercept = threshold_value_cor, colour = "red") +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))+
  theme_bw()

@@

## Coefficient of variation and Variation over time

@@

@@

Following the previous studies of [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @mesoudi2008cultural; @garvey2018current], we only compared CV values of attributes, body length, maximum length, mid width, and tang width. @fig-plot-cv-four-assemblage shows that CV values for the four assemblages are mostly slightly below `r threshold_value`. Among the assemblages, SYG1_2 (n=7) has higher CV values as well as the widest ranges of confidence intervals, perhaps due to its small sample size, relative to the other assemblages compared here. Overall we see only subtle changes in CV values from Phase 2 to Phase 3 in @fig-plot-cv-four-assemblage. This suggests that the complex directional trends in @fig-plot-cv-over-time might be best interpreted as noise in an overall signal of guided variation, rather than substantial changes in the type of bias dominating cultural transmission.
